---
title: "The Impact of Research Investment on Short-Term GDP Growth"
author: "Jeffery Gao, Alexander Caza, Jakin Chan"
date: "2024-04-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction

Research and development investment is essential for major economies to stay
relevant in a rapidly changing technological landscape. Universities and
colleges are one of the major sources of basic and applied research in almost
all industries. For economists and policymakers, understanding the economic
return on investment of government investment in academic research projects is
imperative for implementing intelligent policy decisions with respect to
spending priorities. This paper seeks to investigate how government expenditure
on academic research impacts GDP growth in Ontario. Specifically, it seeks to
investigate the presence and strength of the correlation between such spending
and economic growth in Ontario on a year to year basis.


## Data and Observations

```{r, message=FALSE, warning=FALSE}
set.seed(0)
library(tidyverse)
library(knitr)

# Read data
funding <- read.csv("research_funding_summary_2004-2024_en.csv")
gdp <- read.csv("gdp_goods.csv")

# -------------------------
# FUNDING DATA
# -------------------------

# Convert dollar values into numeric type
funding$Ontario.Commitment <-
  as.numeric(gsub('\\$|,', '', funding$Ontario.Commitment))

# Get data of interest in funding data set
funding <- funding %>%
  mutate(Year = 2000 + as.integer(str_sub(Approval.Date, start=8))) %>%
  select(c("Year", "Ontario.Commitment")) %>%
  group_by(Year) %>%
  summarize(Ontario.Commitment = sum(Ontario.Commitment))

# -------------------------
# GDP DATA
# -------------------------
gdp <- gdp %>%
  filter(Industry == "Goods-Producing Industries ") %>%
  select(-c("NAICS", "Industry")) %>%
  t() %>%
  as_tibble() %>%
  rownames_to_column(var = "Year") %>%
  mutate(Year = as.integer(Year) + 2006) %>%
  rename(GDP = V1)

# Convert dollar values into numeric type
gdp$GDP <- as.numeric(gsub(',', '', gdp$GDP))

# Create GDP Change column
gdp <- gdp %>%
  mutate(gdp_change = (gdp$GDP - lag(gdp$GDP)) / lag(gdp$GDP) * 100)

# -------------------------
# JOINING DATA SETS BY YEAR
# -------------------------
d <- inner_join(funding, gdp, by = "Year") %>%
  select(Year, Ontario.Commitment, gdp_change) %>%
  filter(!is.na(gdp_change))

```

Our report is based on two data sets collected and published by the Government
of Ontario on their Data Catalogue website. The first is the “Ontario Research
Funding – Summary” data set published by the Ontario Government’s Ministry of
Colleges and Universities. This data set provides an overview of all academic
research projects that received Ontario Government funding between October 2004
and March 2024, and is tracked by the government for a variety of purposes
including: accounting and auditing, enabling public oversight, facilitating
economic research and facilitating inter-government and inter-agency
coordination of research efforts. The second is the “Ontario gross domestic
product (GDP) by Goods-Producing Industries and Services Industries” data set
published by the Ontario Government’s Ministry of Agriculture, Food and Rural
Affairs. This data set provides Ontario’s GDP each year, broken down by industry
for the years from 2007 to 2022, and is tracked by the government for a variety
of purposes including: economic research and analysis, providing government
officials and researchers a concise overview of Ontario’s economic health each
year, and for informing the formulation of public policy based on current
economic conditions.

On the next page, we have included the relevant data that is used for this
report.

\pagebreak

```{r}
# -------------------------
# TABLES
# -------------------------

# Funding Data
kable(funding,
      caption = "Ontario Research Funding – Summary",
      col.names = c("Year", "Ontario Funding"))

```

```{r}
# GDP data
kable(gdp,
      caption = paste("Ontario Gross Domestic Product (GDP) by Goods-Producing",
                      "Industries and Services Industries"),
      col.names = c("Year",
                    "GDP (in millions of chained 2012 dollars)",
                    "GDP Percent Change"))

```


## Analysis of Research Investment vs. GDP

In this report, we will analyze whether there is a linear relation between the
amount of funding provided to academic research projects by Ontario and the GDP
percent change by using a linear regression model in R.

```{r, message=FALSE, fig.dim=c(6, 4.25), fig.align="center"}
# -------------------------
# LINEAR REGRESSION MODEL
# (INCLUDING OUTLIER)
# -------------------------

# Linear regression model
l_model <- lm(gdp_change ~ Ontario.Commitment, data = d)
summary(l_model)

# Graph
ggplot(d, aes(x = Ontario.Commitment, y = gdp_change)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Graph 1: Research Funding vs. GDP Percent Change in Ontario",
       x = "Research Funding", y = "GDP Percent Change")

```

This model and its corresponding graph shows the relationship between the change
in GDP and the research investment by Ontario. As shown above, the regression
line (labelled `Ontario.Commitment`) has a slope of $-3.065 \times 10^{-8}$.
This implies that the investing more money will result in a decrease in GDP,
which is unexpected. The calculated $p$-value for the slope is 0.0125,
which indicated that there is a significant relationship between the two
variables. The $R^2$ value, which is $0.3919$, shows that this model only
adequately describes $39\%$ of the data, which is not the greatest.

However, notice the outlier at the bottom right corner of the graph, which
corresponds to the year of the Great Recession, which was the worst economic
downturn since the Great Depression.

So, a second linear regression model was fitted to the data set with the
outlier removed, and this is the result.

```{r}
# -------------------------
# FILTER OUTLIER
# -------------------------
filt_int <- quantile(d$gdp_change, c(0.05, 1))
dfilt <- d %>% filter(gdp_change >= filt_int[1] & gdp_change <= filt_int[2])

```

```{r, message=FALSE}
# -------------------------
# LINEAR REGRESSION MODEL
# (EXCLUDING OUTLIER)
# -------------------------

# Linear regression model
l_model_filt <- lm(gdp_change ~ Ontario.Commitment, data = dfilt)
summary(l_model_filt)

# Graph
ggplot(dfilt, aes(x = Ontario.Commitment, y = gdp_change)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = paste("Graph 2: Research Funding vs. GDP Percent Change",
                     "in Ontario\n(w/o Outliers)"),
       x = "Research Funding", y = "GDP Percent Change")

```

Notice that the slope of the regression line becomes $-4.542 \times 10^{-9}$,
which is slightly less negative than before. Moreover, the $p$-value of the
slope is now $0.756$, which indicates there is very likely no relationship
between the two variables whatsoever. The $R^2$ value of this model is
$0.008327$, which shows that this model is not explaining much of the data
at all. This is likely due to the low amount of data in the given data set.


## Bootstrapping the Linear Regression Model

As the data set is quite small, the data set is likely to be skewed. To more
accurately predict the linear regression model, we have calculated a
bootstrapped $95\%$ confidence interval for the slope of the model.

This is the $95\%$ confidence interval for the slope of the model when the
outlier is included:

```{r}
# -------------------------
# BOOTSTRAPPING LM MODEL
# -------------------------

# Including outlier
boot_function <- function() {
  boot_d <- d %>% sample_n(nrow(d), replace = TRUE)
  boot_m <- lm(gdp_change ~ Ontario.Commitment, data = boot_d)
  s <- coef(boot_m)[2]
  return(s)
}
boot_slope <- replicate(1000, boot_function())
quantile(boot_slope, c(0.025, 0.975))

```

And this is the confidence interval for the slope when the outlier is not
included:

``` {r}
# Excluding outlier
boot_function_filt <- function() {
  boot_d <- dfilt %>% sample_n(nrow(dfilt), replace = TRUE)
  boot_m <- lm(gdp_change ~ Ontario.Commitment, data = boot_d)
  s <- coef(boot_m)[2]
  return(s)
}
boot_slope_filt <- replicate(1000, boot_function_filt())
quantile(boot_slope_filt, c(0.025, 0.975))

```

In both cases, the value $0$ is included within the $95\%$ confidence interval
for the slope. This means that we **cannot** reject the null hypothesis that
there is no correlation between the amount of funding provided by Ontario and
the percentage change in GDP for the following year.


## Cross-Validating the Linear Regression Model

As described above, our model might not be accurately describing the data.
To verify this, we will cross-validate the linear regression model to find the
mean squared error (MSE).

It was observed that the range of outputs that was being calculated varied a
lot, so we decided to bootstrap $95\%$ confidence intervals for the obtained MSE
values.

This is the $95\%$ confidence interval for the MSE when the outlier is included:

```{r}
# -------------------------
# CROSS VALIDATION LM MODEL
# -------------------------

# We used bootstrapping for the MSE obtained from cross validation
# because the results were very sporadic

# Cross validation (w/ outlier)
cv_func <- function() {
  d_cv <- d %>% mutate(group_ind = sample(c("train", "test"),
                                            size = nrow(d),
                                            prob = c(0.6, 0.4),
                                            replace = TRUE))

  cv_m <- lm(gdp_change ~ Ontario.Commitment,
                data = d_cv %>% filter(group_ind == "train"))

  # MSE of Training Data Set
  y_hat_train <- predict(cv_m)
  mse <- mean((d_cv$gdp_change[d_cv$group_ind == "train"] - y_hat_train) ^ 2)
  return(mse)
}

boot_mse <- replicate(1000, cv_func())
quantile(boot_mse, c(0.025, 0.975))

```

And this is the $95\%$ confidence interval for the MSE when the outlier is not
included:

```{r}
# Cross validation (w/o outlier)
cv_func_filt <- function() {
  d_cv <- d %>% mutate(group_ind = sample(c("train", "test"),
                                            size = nrow(d),
                                            prob = c(0.6, 0.4),
                                            replace = TRUE))

  cv_m <- lm(gdp_change ~ Ontario.Commitment,
                data = d_cv %>% filter(group_ind == "train"))

  # MSE of Test Data Set
  y_hat_test <- predict(cv_m, newdata = d_cv %>% filter(group_ind == "test"))
  mse <- mean((d_cv$gdp_change[d_cv$group_ind == "test"] - y_hat_test) ^ 2)
  return(mse)
}

boot_mse_filt <- replicate(1000, cv_func_filt())
quantile(boot_mse_filt, c(0.025, 0.975))

```

As expected, in both cases, the MSE confidence intervals obtained from models
of both data sets are not that close to $0$, indicating that the models are
not explaining the data set very well.

It is also worth pointing out that the confidence interval for the data set
without the outlier is shifted a bit in the positive direction, and the range is
a lot bigger, which shows that the model performing suboptimally when the
outlier is removed.


## Summary

There does not seem to be any correlation between research and development
funding and GDP growth, and if there is, it is a weak negative correlation. This
suggests that either research spending reduces economic growth, or more likely
in the opinion of the authors, so many other factors play a role in economic
growth that the specific impact provincial research funding has is not
detectable, at least in a data set with not very many data sets, and even in a
far larger data set, the correlation is likely still very weak. It should also
be noted that the reliability of our findings is challenged by the fact that our
data set contains the period of the financial crisis of 2008, which may skew
findings. Further avenues for funding impact investigations may also find more
success in generating actionable recommendations for public expenditure by
investigating the long-term effects of research funding, for example by studying
research expenditure as a percentage of GDP and GDP growth over a time span
measured in decades, with the comparison being between different nations, as
opposed to focusing on the year-to-year impact in one region as the authors of
this study did.


\pagebreak

## Appendix: R Code Used in this Report

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```